---
title: "aquamet Documentation"
date: "`r Sys.Date()`"
format: 
  html: 
    code-fold: show
    code-tools: true
    embed-resources: true
    toc: true
author: 
  name: Karen Blocksom
  email: blocksom.karen@epa.gov
  affiliation: 
    name: U.S. EPA, Office of Research and Development
    city: Corvallis
    state: OR
---

# Introduction

The purpose of this document is to provide guidance on using the *aquamet* R package. This package is used to calculate physical habitat metrics from raw field-collected data for the National Lakes Assessment (NLA) and the National Rivers and Streams Assessment (NRSA). In each of these National Aquatic Resource Surveys (NARS), hundreds to thousands of individual measurements are made at varying scales, depending on the measurement. These measurements are then combined into hundreds of metrics, a relatively small number of which are eventually used in physical habitat indicators. There are at least 10 functions for NLA and 15 functions for NRSA to calculate physical habitat metrics, and some require previously calculated metrics. Because the process is so complex, this document has been created to walk the user through the use of the functions.

You will notice that code blocks are hidden by folding them up in the document, and clicking on the arrow next to `Code` will reveal the code used to produce output. If you hover in the upper right corner of each code block, you can copy the code to the clipboard and run it in your own session of R.

### Installing aquamet

The *aquamet* package is available as a tarball (tar.gz extension), which is much like a zip file that combines all of the files needed to run the package. It can be installed with the code below if needed, but you will need to replace `path_to_file` with the actual path. If you need to install this package, click the arrow next to `Code`, copy the code and paste to your R console, change the path to the location of the aquamet source file on your machine, and then run it.

```{r install-package, eval=FALSE}
devtools::install_local("path_to_file", dependencies = 'Imports', type="source")
```

## Installing other necessary packages

Because this package is not located on CRAN, you may need to install the following packages if you do not already have them installed: plyr, dplyr, reshape2, Hmisc, gtools, stringr, foreach, RUnit, and tidyr. This can be done in one statement, using the code below. If you use RStudio, you can also simply use the Install option in the Packages tab to select all of the packages you need to install from CRAN.

```{r install-oth-pkgs, eval=FALSE}
install.packages(c('plyr', 'dplyr', 'reshape2', 'Hmisc', 'gtools', 'stringr', 
                   'foreach', 'Runit', 'tidyr'))
```

# Getting started

First, we need to load the `aquamet` library. You will also want to ensure you have the `tidyr` package installed and loaded because we will use it to made data wider or longer, depending on our needs. There are also a few other packages that we may use for manipulating the data while preparing it for use with the `aquamet` package.

```{r load-library, echo=TRUE, message=FALSE}
library(tidyr)
library(plyr)
library(dplyr)
library(aquamet)
library(Hmisc)
library(PHABdata)
```

The functions in the *aquamet* package are designed to be flexible, but they do assume certain data characteristics. Importantly, they assume there is a single variable that describes the site-visit, and that it is named `SITE`. Thus, if you have multiple visits to some sites in your data, you will need to create a new variable that combines this information and is named `SITE`.

# NLA Example

There is a data package created to accompany this documentation called `PHABdata`. It contains example physical habitat data following the current data structure for NARS. We will use these example datasets to walk through the steps of calculating metrics. We can take a peek at the NLA physical habitat example data using `head()`. The structure of this dataset already has the variable names of `SITE`, `STATION`, `PARAMETER`, and `VALUE` that are expected and required by the NLA metrics functions. However, the inputs to each function are expected to be subsets of individual parameters in the dataset. Therefore, it is useful to see what the parameter names in the dataset are first.

```{r data-look}

head(nlaPhab)

unique(nlaPhab$PARAMETER)
```

It is obvious that there are large numbers of measurements in this dataset, and to better understand where those parameters came from, we can look at the prefix for many of these parameters. The general association of prefix to data type is:

1)  AM: littoral aquatic macrophytes
2)  BS: littoral bottom substrate
3)  C: riparian canopy
4)  FC: littoral fish cover
5)  GC: riparian ground cover
6)  HI: human influence
7)  SS: shoreline substrate
8)  U: riparian understory

Also note that parameters with the suffix of \_DD are measured in the drawdown zone when one is present, and those without the suffix are measured in the standard littoral or riparian zone, depending on the parameter. Metrics based only on drawdown parameters have \_DD added to the metric name, whereas those based only on the riparian or littoral data standard plots have a suffix of \_RIP or \_LIT, respectively. There are also \_SYN metrics which are synthesized by combining the drawdown and riparian data together and \_SIM metrics which combine drawdown and littoral data. These versions of the metrics are meant to simulate the way data were collected in 2007, which did not include drawdown plots. However, it is not important to understand the details of how these data are combined, because if you include both drawdown and riparian/littoral parameter data when you call functions, the functions will automatically output all three types of metrics. Functions that can produce \_SYN metrics are nlaHumanImpact(), nlaRiparianVegetation(), and the function that can produce \_SIM metrics is nlaFishCover(). If drawdown data were not recorded, the resulting metrics would simply be the \_RIP or \_LIT metrics. For the purposes of calculating physical habitat indicators, the \_SYN and \_LIT versions of metrics are used for data from 2012 onward.

## Calculating metrics

Setting up the data for calculating metrics is much simpler and more straightforward for NLA data than for NRSA data. The NLA data are structured in a single format, with all data having the following main variables: SITE, STATION, PARAMETER, VALUE. Let's start by calculating aquatic macrophyte metrics using `nlaAquaticMacrophytes()`. There are several arguments to this function, with each representing the data for a particular parameter. The expected format of those input data are SITE, STATION, VALUE (the parameter value is assumed). We can use the names of the arguments as the guide to which parameter is expected.

For this function, we need to create datasets for emergent vegetation cover (parameter AM_EMERGENT), floating vegetation cover (parameter AM_FLOATING), submergent vegetation cover (parameter AM_SUBMERGENT), and total vegetation cover (parameter AM_TOTALCOVER). Since the expected input datasets only have the variables SITE, STATION, and VALUE, we also need to drop the PARAMETER variable.

```{r calc-aqmac-1}
unique(nlaPhab$PARAMETER)

emergent <- subset(nlaPhab, PARAMETER=='AM_EMERGENT', select = -PARAMETER)

floating <- subset(nlaPhab, PARAMETER=='AM_FLOATING', select = -PARAMETER)

submergent <- subset(nlaPhab, PARAMETER=='AM_SUBMERGENT', select = -PARAMETER)

totcover <- subset(nlaPhab, PARAMETER=='AM_TOTALCOVER', select = -PARAMETER)
```

There are two other arguments in the `nlaAquaticMacrophytes()` function. The one named dataInformation simply defines the different values found in the data and the weights associated with them, as well as an indicator of presence of vegetation. If using standard NLA data, this argument can be left as the default, so it does not need to be specified. The second additional argument is only used when running tests and can be ignored because its default is FALSE. Now that we have the datasets needed to run the function, we can calculate metrics.

```{r calc-aqmet-2}
aqmacroMets <- nlaAquaticMacrophytes(emergent=emergent, 
                                     floating=floating, 
                                     submergent=submergent, 
                                     totalCover=totcover)

head(aqmacroMets)
```

To look at the full set of metrics calculated, we want to pivot the data into a wide format to see all metrics at once.

```{r calc-aqmet-3}
aqmacroMets.wide <- pivot_wider(aqmacroMets, id_cols = 'SITE',
                                names_from='METRIC', 
                                values_from='VALUE')

names(aqmacroMets.wide)


```

The output is a pretty simple dataset because the drawdown zone is not included in the calculations and there is only one version of each metric. Now we can perform the same preparation for the other NLA functions to get the full set of metrics.

We will need to update values for the parameter ANGLE because NEAR_VERTICAL_UNDERCUT is not a valid value. It must be updated to NEAR_VERTICAL in order to use in the function. Thus, we can update the full dataset for this value first. For certain other inputs, the VALUE variable must be converted to numeric to avoid an error.

```{r calc-bank-mets}

angle <- subset(nlaPhab, PARAMETER=='ANGLE', select = -PARAMETER) |>
  plyr::mutate(VALUE = ifelse(VALUE=='NEAR_VERTICAL_UNDERCUT', 'NEAR_VERTICAL', VALUE))

drawdown <- subset(nlaPhab, PARAMETER=='DRAWDOWN', select = -PARAMETER)

horizontalDistance <- subset(nlaPhab, PARAMETER=='HORIZ_DIST', select=-PARAMETER) |>
  plyr::mutate(VALUE = as.numeric(VALUE))

horizontalDistanceDD <- subset(nlaPhab, PARAMETER=='HORIZ_DIST_DD', select=-PARAMETER) |>
  plyr::mutate(VALUE = as.numeric(VALUE))

verticalHeight = subset(nlaPhab, PARAMETER=='VERT_HEIGHT', select=-PARAMETER) |>
  plyr::mutate(VALUE = as.numeric(VALUE))
                            
verticalHeightDD = subset(nlaPhab, PARAMETER=='VERT_HEIGHT_DD', select=-PARAMETER) |>
  plyr::mutate(VALUE = as.numeric(VALUE))
# nlaBankFeatures
bankMets <- nlaBankFeatures(angle,
                            drawdown,
                            horizontalDistance,
                            horizontalDistanceDD,
                            verticalHeight,
                            verticalHeightDD)

unique(bankMets$METRIC)
                            
```

Now that we can see the general format of inputting data to the function and the outputs, we will run the other metric functions, preparing the input data for each one first.

```{r calc-other-mets-1}
# Bottom Substrate
bedrock <- subset(nlaPhab,PARAMETER=='BS_BEDROCK',select=-PARAMETER)
boulder <- subset(nlaPhab,PARAMETER=='BS_BOULDER',select=-PARAMETER)
color <- subset(nlaPhab,PARAMETER=='BS_COLOR',select=-PARAMETER)
cobble <- subset(nlaPhab,PARAMETER=='BS_COBBLE',select=-PARAMETER)
gravel <- subset(nlaPhab,PARAMETER=='BS_GRAVEL',select=-PARAMETER)
odor <- subset(nlaPhab,PARAMETER=='ODOR',select=-PARAMETER)
organic <- subset(nlaPhab,PARAMETER=='BS_ORGANIC',select=-PARAMETER)
sand <- subset(nlaPhab,PARAMETER=='BS_SAND',select=-PARAMETER)
silt <- subset(nlaPhab,PARAMETER=='BS_SILT',select=-PARAMETER)
wood <- subset(nlaPhab,PARAMETER=='BS_WOOD',select=-PARAMETER)
  
botSubMets <- nlaBottomSubstrate(bedrock,
                                 boulder,
                                 color,  
                                 cobble,
                                 gravel,
                                 odor,
                                 organic,
                                 sand,
                                 silt,
                                 wood)

head(botSubMets)
unique(botSubMets$METRIC)

# Shoreline Substrate
bedrock <- subset(nlaPhab,PARAMETER=='SS_BEDROCK',select=-PARAMETER)
boulder <- subset(nlaPhab,PARAMETER=='SS_BOULDERS',select=-PARAMETER)
color <- subset(nlaPhab,PARAMETER=='SS_COLOR',select=-PARAMETER)
cobble <- subset(nlaPhab,PARAMETER=='SS_COBBLE',select=-PARAMETER)
gravel <- subset(nlaPhab,PARAMETER=='SS_GRAVEL',select=-PARAMETER)
organic <- subset(nlaPhab,PARAMETER=='SS_ORGANIC',select=-PARAMETER)
other <- subset(nlaPhab,PARAMETER=='SS_OTHER',select=-PARAMETER)
sand <- subset(nlaPhab,PARAMETER=='SS_SAND',select=-PARAMETER)
silt <- subset(nlaPhab,PARAMETER=='SS_SILT',select=-PARAMETER)
wood <- subset(nlaPhab,PARAMETER=='SS_WOOD',select=-PARAMETER)

shoreSubMets <- nlaShorelineSubstrate(bedrock,
                                 boulder,
                                 cobble,
                                 gravel,
                                 organic,
                                 other,
                                 sand,
                                 silt,
                                 wood)

head(shoreSubMets)
unique(shoreSubMets$METRIC)

# Fish Cover

  aquatic <- subset(nlaPhab,PARAMETER=='FC_AQUATIC',select=-PARAMETER)
  aquatic_dd <- subset(nlaPhab,PARAMETER=='FC_AQUATIC_DD',select=-PARAMETER)
  boulders <- subset(nlaPhab,PARAMETER=='FC_BOULDERS',select=-PARAMETER)
  boulders_dd <- subset(nlaPhab,PARAMETER=='FC_BOULDERS_DD',select=-PARAMETER)
  brush <- subset(nlaPhab,PARAMETER=='FC_BRUSH',select=-PARAMETER)
  brush_dd <- subset(nlaPhab,PARAMETER=='FC_BRUSH_DD',select=-PARAMETER)
  ledges <- subset(nlaPhab,PARAMETER=='FC_LEDGES',select=-PARAMETER)
  ledges_dd <- subset(nlaPhab,PARAMETER=='FC_LEDGES_DD',select=-PARAMETER)
  livetrees <- subset(nlaPhab,PARAMETER=='FC_LIVETREES',select=-PARAMETER)
  livetrees_dd <- subset(nlaPhab,PARAMETER=='FC_LIVETREES_DD',select=-PARAMETER)
  overhang <- subset(nlaPhab,PARAMETER=='FC_OVERHANG',select=-PARAMETER)
  overhang_dd <- subset(nlaPhab,PARAMETER=='FC_OVERHANG_DD',select=-PARAMETER)
  snags <- subset(nlaPhab,PARAMETER=='FC_SNAGS',select=-PARAMETER)
  snags_dd <- subset(nlaPhab,PARAMETER=='FC_SNAGS_DD',select=-PARAMETER)
  structures <- subset(nlaPhab,PARAMETER=='FC_STRUCTURES',select=-PARAMETER)
  structures_dd <- subset(nlaPhab,PARAMETER=='FC_STRUCTURES_DD',select=-PARAMETER)
  drawdown <- subset(nlaPhab,PARAMETER=='DRAWDOWN',select=-PARAMETER)
  
  # Ensure VALUE is numeric for this particular subset
  horizontalDistance_dd <- subset(nlaPhab,PARAMETER=='HORIZ_DIST_DD',
     select=-PARAMETER)
  horizontalDistance_dd$VALUE <- with(horizontalDistance_dd, as.numeric(VALUE))
  
  # Use defaults for fillinDrawdown, createSyntheticCovers, and coverClassInfo
  # arguments
  fishCovMets <- nlaFishCover(aquatic, aquatic_dd,
                              boulders, boulders_dd,
                              brush, brush_dd, 
                              ledges, ledges_dd, 
                              livetrees, livetrees_dd, 
                              overhang, overhang_dd,
                              snags, snags_dd, 
                              structures, structures_dd, 
                              drawdown, horizontalDistance_dd)
  
  head(fishCovMets)
  unique(fishCovMets$METRIC)
```

Where possible, we can subset the example dataset within the call to the metric function. This is demonstrated below. Sometimes, if a particular result must be converted to numeric, we may have to subset before calling the function.

```{r calc-other-mets-2}
  # Human Influence metrics
  humInflMets <- nlaHumanImpact(
    buildings = subset(nlaPhab,PARAMETER=='HI_BUILDINGS',select=-PARAMETER),
    buildings_dd = subset(nlaPhab,PARAMETER=='HI_BUILDINGS_DD',select=-PARAMETER),
    commercial = subset(nlaPhab,PARAMETER=='HI_COMMERCIAL',select=-PARAMETER),
    commercial_dd = subset(nlaPhab,PARAMETER=='HI_COMMERCIAL_DD',select=-PARAMETER),
    crops = subset(nlaPhab,PARAMETER=='HI_CROPS',select=-PARAMETER),
    crops_dd = subset(nlaPhab,PARAMETER=='HI_CROPS_DD',select=-PARAMETER),
    docks = subset(nlaPhab,PARAMETER=='HI_DOCKS',select=-PARAMETER),
    docks_dd = subset(nlaPhab,PARAMETER=='HI_DOCKS_DD',select=-PARAMETER),
    landfill = subset(nlaPhab,PARAMETER=='HI_LANDFILL',select=-PARAMETER),
    landfill_dd = subset(nlaPhab,PARAMETER=='HI_LANDFILL_DD',select=-PARAMETER),
    lawn = subset(nlaPhab,PARAMETER=='HI_LAWN',select=-PARAMETER),
    lawn_dd = subset(nlaPhab,PARAMETER=='HI_LAWN_DD',select=-PARAMETER),
    orchard = subset(nlaPhab,PARAMETER=='HI_ORCHARD',select=-PARAMETER),
    orchard_dd = subset(nlaPhab,PARAMETER=='HI_ORCHARD_DD',select=-PARAMETER),
    other = subset(nlaPhab,PARAMETER=='HI_OTHER',select=-PARAMETER),
    other_dd = subset(nlaPhab,PARAMETER=='HI_OTHER_DD',select=-PARAMETER),
    park = subset(nlaPhab,PARAMETER=='HI_PARK',select=-PARAMETER),
    park_dd = subset(nlaPhab,PARAMETER=='HI_PARK_DD',select=-PARAMETER),
    pasture = subset(nlaPhab,PARAMETER=='HI_PASTURE',select=-PARAMETER),
    pasture_dd = subset(nlaPhab,PARAMETER=='HI_PASTURE_DD',select=-PARAMETER),
    powerlines = subset(nlaPhab,PARAMETER=='HI_POWERLINES',select=-PARAMETER),
    powerlines_dd = subset(nlaPhab,PARAMETER=='HI_POWERLINES_DD',select=-PARAMETER),
    roads = subset(nlaPhab,PARAMETER=='HI_ROADS',select=-PARAMETER),
    roads_dd = subset(nlaPhab,PARAMETER=='HI_ROADS_DD',select=-PARAMETER),
    walls = subset(nlaPhab,PARAMETER=='HI_WALLS',select=-PARAMETER),
    walls_dd = subset(nlaPhab,PARAMETER=='HI_WALLS_DD',select=-PARAMETER),
    drawdown = subset(nlaPhab,PARAMETER=='DRAWDOWN',select=-PARAMETER),
    horizontalDistance_dd = subset(nlaPhab,PARAMETER=='HORIZ_DIST_DD',select=-PARAMETER)
  )
  
head(humInflMets)
unique(humInflMets$METRIC)


# Littoral Zone metrics
litZoneMets <- nlaLittoralZone(littoralFilm = subset(nlaPhab, 
                                PARAMETER=='SURFACE_FILM',select=-PARAMETER), 
                               data2007=FALSE)

head(litZoneMets)
unique(litZoneMets$METRIC)

# Riparian vegetation metrics
ripVegMets <- nlaRiparianVegetation(
  bigTrees = subset(nlaPhab,PARAMETER=='C_BIGTREES',select=-PARAMETER),
  bigTrees_dd = subset(nlaPhab,PARAMETER=='C_BIGTREES_DD',select=-PARAMETER),
  smallTrees = subset(nlaPhab,PARAMETER=='C_SMALLTREES',select=-PARAMETER),
  smallTrees_dd = subset(nlaPhab,PARAMETER=='C_SMALLTREES_DD',select=-PARAMETER),
  canopyType = subset(nlaPhab,PARAMETER=='CANOPY',select=-PARAMETER),
  canopyType_dd = subset(nlaPhab,PARAMETER=='CANOPY_DD',select=-PARAMETER),
  groundcoverBare = subset(nlaPhab,PARAMETER=='GC_BARE',select=-PARAMETER),
  groundcoverBare_dd = subset(nlaPhab,PARAMETER=='GC_BARE_DD',select=-PARAMETER),
  groundcoverInundated = subset(nlaPhab,PARAMETER=='GC_INUNDATED',select=-PARAMETER),
  groundcoverInundated_dd = subset(nlaPhab,PARAMETER=='GC_INUNDATED_DD',select=-PARAMETER),
  groundcoverNonwoody = subset(nlaPhab,PARAMETER=='GC_NONWOODY',select=-PARAMETER),
  groundcoverNonwoody_dd = subset(nlaPhab,PARAMETER=='GC_NONWOODY_DD',select=-PARAMETER),
  groundcoverWoody = subset(nlaPhab,PARAMETER=='GC_WOODY',select=-PARAMETER),
  groundcoverWoody_dd = subset(nlaPhab,PARAMETER=='GC_WOODY_DD',select=-PARAMETER),
  understoryNonwoody = subset(nlaPhab,PARAMETER=='U_NONWOODY',select=-PARAMETER),
  understoryNonwoody_dd = subset(nlaPhab,PARAMETER=='U_NONWOODY_DD',select=-PARAMETER),
  understoryWoody = subset(nlaPhab,PARAMETER=='U_WOODY',select=-PARAMETER),
  understoryWoody_dd = subset(nlaPhab,PARAMETER=='U_WOODY_DD',select=-PARAMETER),
  understoryType = subset(nlaPhab,PARAMETER=='UNDERSTORY',select=-PARAMETER),
  understoryType_dd = subset(nlaPhab,PARAMETER=='UNDERSTORY_DD',select=-PARAMETER),
  drawdown = subset(nlaPhab,PARAMETER=='DRAWDOWN',select=-PARAMETER),
  horizontalDistance_dd = subset(nlaPhab,PARAMETER=='HORIZ_DIST_DD',select=-PARAMETER),
)

unique(ripVegMets$METRIC)

# NLA station information
isIsland <- subset(nlaPhab,PARAMETER=='ISLAND' & VALUE %in% c('YES','Y'),
  select=-PARAMETER)
  
depth <- subset(nlaPhab,PARAMETER=='DEPTH_AT_STATION',select=-PARAMETER)
depth$VALUE <- as.numeric(depth$VALUE)
  
statInfoMets <- nlaStationInformation(isIsland = isIsland, stationDepth = depth)

unique(statInfoMets$METRIC)

```

Now we will want to combine all of the metrics into a single data frame and pivoting them wider for ease of use later.

```{r combine-metrics}
nlaMets <- rbind(aqmacroMets, bankMets, botSubMets, shoreSubMets,
                 fishCovMets, humInflMets, litZoneMets, 
                 ripVegMets, statInfoMets) 

mets_to_numeric <- subset(nlaMets, METRIC %nin% c('BFOANGLE', 'BSOPCLASS', 'BSOFCLASS', 'BSOCOLOR', 'BSOODOR', 'SSOPCLASS', 'SSOFCLASS', 'LZOFILM'), select = METRIC) |>
  unique()

nlaMets.wide <- pivot_wider(nlaMets, id_cols = 'SITE', names_from='METRIC', values_from='VALUE') |>
  mutate(across(mets_to_numeric$METRIC, as.numeric))
```

## Calculating indicators

Now that all of the metrics have been calculated, we have much of the information needed to calculate the NLA physical habitat indicators and assign condition classes to each site. There are 5 indicators: NLA Drawdown Indicator (`nlaDrawdownIndicator()`), NLA Riparian Vegetation Complexity Indicator (`nlaRipVegCompIndicator()`), NLA Littoral Vegetation Complexity Indicator (`nlaLitVegCompIndicator()`), NLA Littoral-Riparian Vegetation Complexity Indicator (`nlaLitRipVegCompIndicator`), and the NLA Lakeshore Anthropogenic Disturbance Indicator (`nlaRipDistIndicator()`). For each of these indicators, we may require additional information about each site, such as latitude, longitude, lake area, lake elevation, ecoregion, and lake origin. This is because all of the physical habitat indicators except the NLA Lakeshore Anthropogenic Disturbance Indicator compare the observed value for a particular indicator with a modeled site-specific expectation. Typically, the site-related data would be obtained using GIS, but for our purposes, we will create example data in our code to provide all of the additional variables necessary to assign condition. The latitude and longitude represent the approximate center of the lake, and AG_ECO9 is the aggregated level 3 Omernik ecoregion where the center of the lake is located. The lake origin is assigned as either NATURAL or MAN_MADE, the lake area is in square kilometers, and the elevation is the lake surface elevation in meters above sea level.

```{r create-site-variables}
sitedata <- data.frame(SITE=c(6400, 6469, 6768, 6865, 6869, 7623, 
                              8184, 8251, 8657, 1000048),
                       AG_ECO9=c('SAP', 'SAP', 'TPL', 'NPL', 'XER',
                                 'WMT', 'WMT', 'WMT', 'WMT','CPL'),
                       LAKE_ORIGIN=c('MAN_MADE', 'MAN_MADE',
                                     'MAN_MADE', 'NATURAL', 
                                     'MAN_MADE', 'NATURAL', 
                                     'MAN_MADE', 'MAN_MADE', 
                                     'NATURAL', 'MAN_MADE'), 
                       LAT_DD=c(35.72678, 36.12929, 41.46944, 
                                48.00697, 38.84754, 45.86918, 
                                40.05559, 34.03126, 48.568425306, 
                                33.067121087),
                       LON_DD=c(-82.08422, -79.83690, -93.92054, 
                                -101.53242, -111.96139, -113.54893,
                                -105.74708, -109.44311, 
                                -123.0735207, -95.73801377),
                       ELEV=c(427.50, 251.63, 278.75, 620.39, 
                              1589.51, 2412.81, 3029.04, 2519.09,
                              47.75, 148.99),
                       AREA=c(0.69732608, 0.02485657, 0.93282221, 
                              0.74152161, 0.93174515, 0.07763512, 
                              0.51027966, 0.17282305, 0.26395493, 
                              0.14714400))

sitedata

indicIn <- merge(sitedata, nlaMets.wide, by = 'SITE')

```

### NLA Drawdown Indicator

We will start with the drawdown indicator, which measures the lake drawdown exposure. This function requires the aggregated ecoregion and the lake origin as inputs, as well as metrics related to horizontal and vertical drawdown: mean vertical drawdown (BFXVERTHEIGHT_DD) and mean horizontal drawdown (BFXHORIZDIST_DD), both from the `nlaBankFeatures` function. The `nlaDrawdownIndicator` function typically provides 3 output variables, one assigning a class to vertical drawdown and one for horizontal drawdown, then one for overall drawdown condition (`DRAWDOWN_COND`). Each of these assigns a class with possible values of Small, Medium, or Large. For only 2017 data, there is a different indicator used that only assigns values of Not Large or Large, due to data collection issues.

```{r drawdown}
drawdownCond <- nlaDrawdownIndicator(indicIn, sampID='SITE', 
                                     bfxVertDD='BFXVERTHEIGHT_DD', 
                                     bfxHorizDD='BFXHORIZDIST_DD', 
                                     ecoreg='AG_ECO9', 
                                     lake_origin='LAKE_ORIGIN')

drawdownCond
```

### NLA Lakeshore Anthropogenic Disturbance Indicator

This is the only indicator that is not site-specific. It is based on the metrics from the `nlaHumanImpact` function, specifically those synthesized from the standard and drawdown plots to simulate the NLA 2007 approach. These are metrics ending in SYN representing the index of agricultural influences (HIIAG_SYN), the index of non-agricultural influences (HIINONAGSYN), and the fractional presence of any human influence in station plots (HIFPANYCIRCA_SYN).

The output of the indicator function is the calculated indicator value (RDis_IX) and the condition class assignment (RDIS_COND), with values of Good/Fair/Poor.

```{r rip-dist}
ripdistCond <- nlaRipDistIndicator(indicIn, sampID = 'SITE',
                                   hiiAg = 'HIIAG_SYN', 
                                   hiiNonAg = 'HIINONAG_SYN',
                                   hifpAnyCirca = 'HIFPANYCIRCA_SYN')

ripdistCond 
```

### NLA Riparian Vegetation Complexity Indicator

This indicator measures the riparian vegetation cover on the lakeshore. The function requires a large number of site-related variables, including: latitude, longitude, lake origin, lake area, elevation, and ecoregion. It also requires a number of physical habitat metrics to calculate the value used to assign condition. These include riparian vegetation, shoreline substrate, and human impact metrics, listed below with descriptions. The metrics come from the `nlaRiparianVegetation()`, `nlaShorelineSubstrate()`, and `nlaHumanImpact()` functions.

| Argument         | Metric               | Description                                                                     |
|-------------------|----------------------|-------------------------------|
| rviWoody         | RVI_WOODY_SYN        | Index of total woody vegetations cover, synthesizing the 2007 protocol          |
| rvfcGndInundated | RVFCGNDINUNDATED_SYN | Fraction of ground cover by inundation synthesizing the 2007 protocol           |
| rvfcUndWoody     | RVFCUNDWOODY_SYN     | Fraction of understory cover by woody vegetation synthesizing the 2007 protocol |
| rvfcGndWoody     | RVFCGNDWOODY_SYN     | Fraction of ground cover by woody vegetation synthesizing the 2007 protocol     |
| rvfpCanBig       | RVFPCANBIG_SYN       | Fraction of canopy with large trees present synthesizing the 2007 protocol      |
| ssfcBedrock      | SSFCBEDROCK          | Fractional cover of bedrock using SS_BEDROCK                                    |
| ssfcBoulders     | SSFCBOULDERS         | Fractional cover of bedrock using SS_BOULDER                                    |
| hipwWalls        | HIPWWALLS_SYN        | Weighted presence of Walls synthesizing the 2007 protocol                       |

```{r ripveg-comp}
ripvegOut <- nlaRipVegCompIndicator(indicIn, 
                                    sampID='SITE', 
                                    lat='LAT_DD', 
                                    lon='LON_DD',
                                    lake_origin='LAKE_ORIGIN',
                                    area='AREA', 
                                    elev='ELEV', 
                                    ecoreg='AG_ECO9', 
                                    rviWoody='RVIWOODY_SYN',
                                    rvfcGndInundated = 'RVFCGNDINUNDATED_SYN', 
                                    rvfcUndWoody='RVFCUNDWOODY_SYN', 
                                    rvfcGndWoody='RVFCGNDWOODY_SYN',
                                    rvfpCanBig='RVFPCANBIG_SYN',
                                    ssfcBedrock='SSFCBEDROCK', 
                                    ssfcBoulders='SSFCBOULDERS', 
                                    hipwWalls='HIPWWALLS_SYN')

ripvegOut
```

The output of this function provides three variables along with the condition variable. RVegQ represents the observed riparian vegetation complexity and RVegQc3x15 is the site-specific expectation for riparian complexity, based site-related input variables. RVegQc3OE is the observed/expected ratio used to assign the condition class.

These results show that when any critical metrics are missing, Not Assessed is assigned as the condition. In this example, SITE 8251 is missing a few important variables for riparian vegetation.

### NLA Littoral Vegetation Complexity Indicator

This indicator measures the shallow water habitat of the lake. This function also requires latitude, longitude, lake origin, lake area, lake elevation, and ecoregion. The 9 metrics listed below, related to fish cover and aquatic macrophytes, are required as inputs for calculation of this indicator. These metrics come from the `nlaAquaticMacrophytes()` and `nlaFishCover()` functions.

| Argument      | Metric            | Description                                                        |
|-------------------|--------------------|---------------------------------|
| fciNatural    | FCINATURAL_LIT    | Index of fish cover due to natural structures in the littoral zone |
| fcfcSnag      | FCFCSNAGS_LIT     | Fractional cover of snags in the littoral zone                     |
| amfcFloating  | AMFCFLOATING      | Fractional cover of floating macrophytes                           |
| amfcEmergent  | AMFCEMERGENT      | Fractional cover of emergent macrophytes                           |
| fcfcBoulders  | FCFCBOULDERS_LIT  | Fractional cover of boulders in the littoral zone                  |
| fcfcBrush     | FCFCBRUSH_LIT     | Fractional cover of brush in the littoral zone                     |
| fcfcLedges    | FCFCLEDGES_LIT    | Fractional cover of ledges in the littoral zone                    |
| fcfcLiveTrees | FCFCLIVETREES_LIT | Fractional cover of live trees in the littoral zone                |
| fcfcOverhang  | FCFCOVERHANG_LIT  | Fractional cover of overhangs in the littoral zone                 |

```{r litveg-comp}
litvegOut <- nlaLitVegCompIndicator(indicIn, sampID='SITE',
                                    lat='LAT_DD', 
                                    lon='LON_DD', 
                                    lake_origin='LAKE_ORIGIN', 
                                    area='AREA', 
                                    elev='ELEV',
                                    ecoreg='AG_ECO9', 
                                    fciNatural='FCINATURAL_LIT', 
                                    fcfcSnag='FCFCSNAGS_LIT', 
                                    amfcFloating='AMFCFLOATING', 
                                    amfcEmergent='AMFCEMERGENT', 
                                    fcfcBoulders='FCFCBOULDERS_LIT', 
                                    fcfcBrush='FCFCBRUSH_LIT', 
                                    fcfcLedges='FCFCLEDGES_LIT', 
                                    fcfcLiveTrees='FCFCLIVETREES_LIT', 
                                    fcfcOverhang='FCFCOVERHANG_LIT')

litvegOut
```

Similarly to the riparian vegetation complexity indicator output, the observed littoral vegetation complexity (LitCvrQ), the site-specific expected value (LitCvrQc3x15), and the observed/expected ratio (LitCvrQc3OE) are all provided by the \`nlaLitVegCompIndicator\` function. The LITCVR_COND represents the condition class assigned based on the latter value.

Also, there is one site with the assignment of 'Not Assessed'. This is because there were no aquatic macrophyte data collected for that site.

### NLA Littoral-Riparian Vegetation Complexity Indicator

Finally, we will use the results from the littoral and riparian vegetation complexity indicators to calculate indicator values and condition classes for the combined indicator of overall lake habitat complexity. This function requires latitude, longitude, lake origin, lake area, lake elevation, aggregated ecoregion, and the two variables for observation littoral and riparian vegetation complexity, `LitCvrQ` and `RVegQ`.

Note that there are cases where the condition class is 'Not Assessed'. This is because one site was missing the riparian vegetation complexity indicator and one the littoral complexity indicator.

```{r lit-rip-comp}
litripIn <- merge(sitedata, litvegOut, by='SITE') |>
  merge(ripvegOut, by='SITE')

litripOut <- nlaLitRipVegCompIndicator(litripIn,
                                       sampID='SITE',
                                       lat='LAT_DD',
                                       lon='LON_DD',
                                       lake_origin='LAKE_ORIGIN',
                                       area='AREA',
                                       elev='ELEV',
                                       ecoreg='AG_ECO9', 
                                       rvegq='RVegQ',
                                       litcvrq='LitCvrQ')

litripOut
```

Now we have calculated all of the NLA physical habitat indicators, going from raw data to the final classes used in the assessment.

# NRSA Example

Now that we have gone through a simpler example with NLA, we can move on to NRSA, which has a much more complicated data structure. The reason for the increased complexity is two-fold: the NRSA has different field methods for wadeable and boatable sites, and there are various scales at which data are collected in each (e.g., transect level, stations between transects, and along the banks). The data are thus stored in separate tables, depending on the data structure, unlike the NLA data, which can be stored in a single table.

As we go through this example, we will explain for each metric calculation function which type of system the data would come from and the type of data (based on field forms). For the NRSA example, there are a number of different example datasets, which are separated by data type and structure. The datasets are described below.

## Example datasets

| Dataset       | Description                                                                                                                                                                                                 |
|--------------------|----------------------------------------------------|
| nrsaChanconst | Channel constraint characteristics for entire site.                                                                                                                                                         |
| nrsaChannel   | Channel characteristics for both boatable and wadeable streams by TRANSECT, including data on bank characteristics, fish cover, large woody debris, littoral substrates, and boatable constraint variables. |
| nrsaChanrip   | Channel riparian data by TRANSECT and BANK. Includes canopy cover densiometer readings, human influence, visual riparian measurements, and wadeable bank undercut and angle measurements.                   |
| nrsaChanxsec  | Channel cross-section measurements by TRANSECT and BANK for wadeable sites.                                                                                                                                 |
| nrsaInvleg    | Riparian legacy and invasive tree measurements (no longer used).                                                                                                                                            |
| nrsaLittoral  | Nonwadeable littoral depth measurements by TRANSECT and LINE.                                                                                                                                               |
| nrsaSlope     | Slope and bearing measurements.                                                                                                                                                                             |
| nrsaThalweg   | Thalweg depth measurements by TRANSECT and STATION.                                                                                                                                                         |
| nrsaVisits    | Site and type of site (VALXSITE), as BOATABLE or WADEABLE.                                                                                                                                                  |

We can take a quick look at each dataset to see what the data look like.

### nrsaChanconst

```{r nrsa-chanconst}
head(nrsaChanconst)
str(nrsaChanconst)
```

### nrsaChannel

This data frame contains a large number of different types of data, so it is useful to look at all of the combinations of sample types and parameters.

```{r nrsa-channel}
head(nrsaChannel)
str(nrsaChannel)

unique(nrsaChannel[,c('SAMPLE_TYPE', 'PARAMETER')]) |>
  arrange(SAMPLE_TYPE, PARAMETER)
```

### nrsaChanrip

```{r nrsa-chanrip}
head(nrsaChanrip)
str(nrsaChanrip)

unique(nrsaChanrip[,c('SAMPLE_TYPE', 'PARAMETER')]) |>
  arrange(SAMPLE_TYPE, PARAMETER)
```

### nrsaChanxsec

This is for wadeable sites only.

```{r nrsa-chanxsec}
head(nrsaChanxsec)
str(nrsaChanxsec)

unique(nrsaChanxsec[,c('SAMPLE_TYPE', 'PARAMETER')]) |>
  arrange(SAMPLE_TYPE, PARAMETER)
```

### nrsaInvleg

We will show calculations using these data, but note that these measurements are no longer collected, though related but simplified measurements are still collected on the Site Assessment form.

```{r nrsa-invleg}
head(nrsaInvleg)
str(nrsaInvleg) 

unique(nrsaInvleg$PARAMETER)

```

### nrsaLittoral

These are littoral depth measurements only collected in boatable sites along the shoreline.

```{r nrsa-littoral}
head(nrsaLittoral)
str(nrsaLittoral) 

unique(nrsaLittoral$PARAMETER)
```

### nrsaSlope

These are slope and bearing measurements. These are typically only measured in the field for wadeable sites and calculated for boatable sites using GIS. In our example data from the NRSA 2008-9 survey, however, field crews measured slope and bearing at both boatable and wadeable sites.

```{r nrsa-slope}
head(nrsaSlope)
str(nrsaSlope)

unique(nrsaSlope$PARAMETER)
```

### nrsaThalweg

Thalweg depth measurements for both boatable and wadeable sites, with different measurements in each type of site.

```{r nrsa-thalweg}
head(nrsaThalweg)
str(nrsaThalweg)

unique(nrsaThalweg[,c('SAMPLE_TYPE', 'PARAMETER')]) |>
  arrange(SAMPLE_TYPE, PARAMETER)
```

### nrsaVisits

This is just a simple data frame to identify sites as wadeable or boatable.

```{r nrsaVisits}
nrsaVisits
```

## Calculating metrics

### NRSA bank morphology metrics

These are calculated using the `nrsaBankMorphology()` function, and data are provided separately for boatable and wadeable sites. To avoid creating lots of smaller datasets just to create inputs for the metric functions, we can do subsetting and select variables within the function call, using the pipeline function (`|>`) to carry out multiple steps. This approach is shown in the function call below and throughout. 

```{r nrsaBankMorph}
chanrip <- filter(nrsaChanrip, PARAMETER %in% c('ANGLE','UNDERCUT')) 
chan <- filter(nrsaChannel, PARAMETER == 'BKANGLE')

bankMorphMets <- nrsaBankMorphology(bAngle = select(chan, SITE, VALUE), 
                        wAngle = subset(chanrip, PARAMETER=='ANGLE') |> 
                          mutate(VALUE=as.numeric(VALUE)) |>
                          select(SITE, VALUE),
                        wUndercut = subset(chanrip, PARAMETER=='UNDERCUT') |>
                          mutate(VALUE=as.numeric(VALUE)) |>
                          select(SITE, VALUE)
                        )

print("The bank morphology metrics from this function:")
unique(bankMorphMets$METRIC)

```

### NRSA canopy densiometer metrics

These metrics are calculated using the `nrsaCanopyDensiometer()` function, and all data come from the `nrsaChanrip` dataset.

```{r nrsaCanopyDensio}
canDensMets <- nrsaCanopyDensiometer(bDensiom = nrsaChanrip |> 
                                     subset(SAMPLE_TYPE=='CANCOVERB' & PARAMETER=='DENSIOM') |> 
                                     dplyr::rename(DIRECTION=BANK) |> 
                                     subset(VALUE %nin% c('',NA,'NA')) |> 
                                     mutate(VALUE=as.numeric(VALUE)) |>
                                     select(SITE, VALUE),
                           wDensiom = nrsaChanrip |> 
                                     subset(SAMPLE_TYPE=='CANCOVERW' & PARAMETER=='DENSIOM') |>
                                     dplyr::rename(DIRECTION=BANK) |> 
                                     subset(VALUE %nin% c('',NA,'NA')) |> 
                                     mutate(VALUE=as.numeric(VALUE)) |>
                                     select(SITE, DIRECTION, VALUE)
                           )

print("The canopy density metrics from this function:")
unique(canDensMets$METRIC)
```

### NRSA channel characteristics

Channel characteristics metrics are calculated by the `nrsaChannelChar()` function.

```{r nrsaChannelChar}
chanCharMets <- nrsaChannelChar(
                     bankfullWidth = nrsaChanconst |> 
                       subset(PARAMETER == 'BANKFULL') |> 
                       select(SITE, VALUE) |> 
                       mutate(VALUE=as.numeric(VALUE)),
                     channelPattern = nrsaChanconst |> 
                       subset(PARAMETER == 'PATTERN') |>
                       select(SITE, VALUE),
                     constraintFeatures = nrsaChanconst |>
                       subset(PARAMETER == 'FEATURES') |> 
                       select(SITE, VALUE),
                     constraintMultiple = nrsaChannel |> 
                       subset(PARAMETER == 'CONSTRT') |>
                       select(SITE, VALUE),
                     constraintSingle = nrsaChanconst |> 
                       subset(PARAMETER == 'CONSTRNT') |>
                       select(SITE, VALUE),
                     constraintPercent = nrsaChanconst |> 
                       subset(PARAMETER == 'PERCENT') |>
                       select(SITE, VALUE) |>
                       mutate(VALUE=as.numeric(VALUE)),
                     seeOverBank = nrsaChannel |>
                       subset(PARAMETER == 'SEEOVRBK') |> 
                       select(SITE, VALUE),
                     shoreToVegDistance = nrsaChannel |>
                       subset(PARAMETER == 'SHOR2RIP') |>
                       select(SITE, VALUE) |>
                       mutate(VALUE=as.numeric(VALUE)),
                     valleyConstraintUnseen = nrsaChanconst |> 
                       subset(PARAMETER == 'VALLEYBOX') |>
                       select(SITE, VALUE),
                     valleyWidth = nrsaChanconst |> 
                       subset(PARAMETER == 'VALLEY') |>
                       select(SITE, VALUE) |>
                       mutate(VALUE=as.numeric(VALUE))
                     )

print("The channel characteristics metrics from this function:")
unique(chanCharMets$METRIC)
```

### NRSA channel habitat metrics

To calculate channel habitat metrics, which are based on thalweg depth measurements, use the function `nrsaChannelHabitat()`. We need to separate the boatable and wadeable data for these calculations.

```{r nrsaChannelHabitat}

chanHabMets <- nrsaChannelHabitat(
  bChannelUnit = nrsaThalweg |> 
    subset(PARAMETER=='CHANUNCD' & SAMPLE_TYPE=='THALB') |> 
    mutate(VALUE=trimws(VALUE)) |>
    select(SITE, VALUE),
  wChannelUnit = nrsaThalweg |> 
    subset(PARAMETER=='CHANUNCD' & SAMPLE_TYPE=='THALW') |>
    mutate(VALUE=ifelse(VALUE=='PO', 'P', trimws(VALUE))) |>
    select(SITE, VALUE))

print("The channel habitat metrics from this function:")
unique(chanHabMets$METRIC)
```

### NRSA channel morphology metrics

To calculate channel morphology metrics, we use the function `nrsaChannelMorphology()`. The wadeable and boatable data are separate inputs to the function.

```{r nrsaChannelMorphology}
thalweg <- nrsaThalweg |>
  filter(PARAMETER %in% c('DEPTH','DEPTH_UNITS','WETWIDTH','POLE','SONAR')) |> 
  mutate(VALUE=trimws(VALUE))

channel <- nrsaChannel |>
  filter(PARAMETER %in% c('BANKWID', 'WETWID', 'INCISED','BANKHT')) |> 
  mutate(VALUE=trimws(VALUE), UNITS='M')

wthalDepth <- nrsaThalweg |> 
              subset(PARAMETER %in% c('DEPTH')) |> 
              select(-PARAMETER) |> 
              mutate(UNITS='CM') |>
              subset(VALUE %nin% c(NA,''))
bthalDepth <- merge(subset(thalweg, PARAMETER %in% c('POLE','SONAR')) |> 
                  select(-PARAMETER),
                  subset(thalweg, PARAMETER %in% c('DEPTH_UNITS')) |> 
                  dplyr::rename(UNITS=VALUE) |>
                  select(-PARAMETER, -STATION), 
                   by=c('SITE','SAMPLE_TYPE','TRANSECT'),
                   all.x=TRUE
                   ) |>
              mutate(UNITS = ifelse(is.na(UNITS), '', UNITS)) 

verification <- nrsaVisits |>
  mutate(VALXSITE=trimws(VALXSITE))

wadedSites <- filter(verification, VALXSITE=='WADEABLE')$SITE
# NOTE: site 138500 is classed as boatable but has wadeable data.  There are 300 sites with NONE protocol, but those don't show up here.

chanMorphMets <- nrsaChannelMorphology(
  bBankHeight = subset(nrsaChannel, SAMPLE_TYPE=='BANKB' & PARAMETER %in% c('BANKHT')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |> 
    select(SITE, TRANSECT, VALUE, UNITS),
  bBankWidth = subset(nrsaChannel, SAMPLE_TYPE=='BANKB' & PARAMETER %in% c('BANKWID')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, VALUE, UNITS),
  bDepth = bthalDepth |> 
    mutate(VALUE=as.numeric(VALUE)) |>
    select(SITE, TRANSECT, STATION, VALUE, UNITS),
  bIncisedHeight = subset(nrsaChannel, SAMPLE_TYPE=='BANKB' & PARAMETER %in% c('INCISED')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |> 
    select(SITE, TRANSECT, VALUE, UNITS),
  bWettedWidth = subset(nrsaChannel, SAMPLE_TYPE=='BANKB' & PARAMETER %in% c('WETWID')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, VALUE, UNITS),
  wBankHeight = subset(nrsaChannel, SAMPLE_TYPE=='BANKW' & PARAMETER %in% c('BANKHT')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, VALUE, UNITS),
  wBankWidth = subset(nrsaChannel, SAMPLE_TYPE=='BANKW' & PARAMETER %in% c('BANKWID')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, VALUE, UNITS),
  wDepth = wthalDepth |> 
    mutate(VALUE=as.numeric(VALUE)) |> 
    select(SITE, TRANSECT, STATION, VALUE, UNITS),
  wIncisedHeight =subset(nrsaChannel, SAMPLE_TYPE=='BANKW' & PARAMETER %in% c('INCISED')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, VALUE, UNITS),
  wWettedWidth = subset(nrsaThalweg, SAMPLE_TYPE=='THALW' & PARAMETER %in% c('WETWIDTH')) |>
    mutate(VALUE=as.numeric(VALUE), UNITS='M') |>
    select(SITE, TRANSECT, STATION, VALUE, UNITS)
  )

print("The channel morphology metrics from this function:")
unique(chanMorphMets$METRIC)
```

### NRSA fish cover metrics

Unsurprisingly at this point, fish cover metrics are calculated using the function `nrsaFishCover()`. The data for these metrics are located in the example dataset `nrsaChannel`.

```{r nrsaFishCover}
fishCovMets <- nrsaFishCover(
  algae = nrsaChannel |> 
    subset(PARAMETER=='ALGAE') |>
    select(SITE, TRANSECT, VALUE),
  boulder = nrsaChannel |> 
    subset(PARAMETER=='BOULDR') |> 
    select(SITE, TRANSECT, VALUE),
  brush = nrsaChannel |> 
    subset(PARAMETER=='BRUSH') |> 
    select(SITE, TRANSECT, VALUE),
  liveTree = nrsaChannel |> 
    subset(PARAMETER=='LVTREE') |> 
    select(SITE, TRANSECT, VALUE),
  macrophytes = nrsaChannel |> 
    subset(PARAMETER=='MACPHY') |> 
    select(SITE, TRANSECT, VALUE),
  overhang = nrsaChannel |> 
    subset(PARAMETER=='OVRHNG') |> 
    select(SITE, TRANSECT, VALUE),
  structures = nrsaChannel |> 
    subset(PARAMETER=='STRUCT') |> 
    select(SITE, TRANSECT, VALUE),
  undercut = nrsaChannel |> 
    subset(PARAMETER == 'UNDCUT') |> 
    select(SITE, TRANSECT, VALUE),
  woodyDebris = nrsaChannel |> 
    subset(PARAMETER=='WOODY') |> 
    select(SITE, TRANSECT, VALUE)
  )

print("The fish cover metrics from this function:")
unique(fishCovMets$METRIC)
```

### NRSA general site metrics

There is a set of general site metrics that are calculated as part of NRSA physical habitat: reach length, mean transect spacing, number of side channel transects in reach, and % of reach with side channels. Those are calculated using the function `nrsaGeneral()`. A large part of the preparation for this function involves creating data frames of the transect spacing, which differs between boatable and wadeable sites.

#### Boatable transect spacing

In straightforward cases, we can just use the ACTRANSP variable from channel data (e.g., `nrsaChannel`) boatable sites to calculate transect spacing. However, there is another approach using the latitude and longitude measured at the top of each subreach (e.g., transect A to B, B to C, etc.). This may be a more accurate measure of transect spacing in larger boatable systems. For the purposes of providing a tool to use with coordinates, we will follow this approach. First, we must add a function (`greatCircleDistance()`) that calculates the distance between transects using latitude and longitude. We also add a second function (`protectedGreatCircleDistance`) that examines data while calculating values to avoid errors with the `greatCircleDistance()` function.

```{r nrsa-grcirclfx}
greatCircleDistance <- function(lat1, lon1, lat2, lon2)
# Performs 'great circle' distance calculation between two lat/lon points using
# the Vincenty method and WGS-84 geodesic approximations.
#
# Returns distance in meters (NA if no convergence), or a character string
# describing the error if one occurs.
#
# Initially from code by Mario Pineda-Koch described at
# http://www.r-bloggers.com/great-circle-distance-calculations-in-r/  and
# http://pineda-krch.com/2010/11/23/great-circle-distance-calculations-in-r/
#
# modified as follows:
#   a) If the while loop is not entered, it is assumed that the distance between
#      the lat/lon pairs is zero, so 0 is returned
#   b) A clause was added (iterLimit==100) was added to insure that the while
#      loop is entered.  This obviates need for change (a).
#
# ARGUMENTS:
# lat1      latitude in decimal degrees for one point, North is positive
# lon1      longitude in decimal degrees for one point, East is positive
# lat2      latitude in decimal degrees for other point, North is positive
# lon2      longitude in decimal degrees for other point, East is positive
#
{
  # WGS-84 ellipsoid parameters
  a <- 6378137         # length of major axis of the ellipsoid (radius at equator)
  b <- 6356752.314245  # length of minor axis of the ellipsoid (radius at the poles)
  f <- 1/298.257223563 # flattening of the ellipsoid

  # Convert angles in degrees to radians, as expected by the trig functions
  lat1 <- lat1*2*pi/360
  lon1 <- lon1*2*pi/360
  lat2 <- lat2*2*pi/360
  lon2 <- lon2*2*pi/360

  # Pre-calculate frequently used values
  L <- lon2-lon1                 # difference in longitude
  U1 <- atan((1-f) * tan(lat1))  # reduced latitude
  U2 <- atan((1-f) * tan(lat2))  # reduced latitude
  sinU1 <- sin(U1)
  cosU1 <- cos(U1)
  sinU2 <- sin(U2)
  cosU2 <- cos(U2)

  # Initialize values and attempt convergence.
  cosSqAlpha <- NULL
  sinSigma <- NULL
  cosSigma <- NULL
  cos2SigmaM <- NULL
  sigma <- NULL
  lambda <- L
  lambdaP <- 0
  iterLimit <- 100

  while (iterLimit==100 | (abs(lambda-lambdaP) > 1e-12 & iterLimit>0) ) {
      sinLambda <- sin(lambda)
      cosLambda <- cos(lambda)
      sinSigma <- sqrt( (cosU2 * sinLambda) * (cosU2 * sinLambda) +
                        (cosU1 * sinU2 - sinU1 * cosU2 * cosLambda) *
                        (cosU1 * sinU2 - sinU1 * cosU2 * cosLambda)
                      )

      if (sinSigma==0) return(0)  # Co-incident points
      
      cosSigma <- sinU1 * sinU2 + cosU1 * cosU2 * cosLambda
      sigma <- atan2(sinSigma, cosSigma)
      sinAlpha <- cosU1 * cosU2 * sinLambda / sinSigma
      cosSqAlpha <- 1 - sinAlpha * sinAlpha
      cos2SigmaM <- cosSigma - 2 * sinU1 * sinU2 / cosSqAlpha
      if (is.na(cos2SigmaM)) cos2SigmaM <- 0  # Equatorial line: cosSqAlpha=0
      
      C <- f/16 * cosSqAlpha * (4 + f * (4 - 3 * cosSqAlpha))

      lambdaP <- lambda
      lambda <- L + (1 - C) * f * sinAlpha *
                (sigma + C * sinSigma *
                 (cos2SigmaM + C * cosSigma * (-1 + 2 * cos2SigmaM * cos2SigmaM))
                )
      iterLimit <- iterLimit - 1
  }

  if (iterLimit==0) return(NA)  # formula failed to converge
  if (iterLimit==100) return(0)  # formula began with nearly or exactly coincident points

  uSq <- cosSqAlpha * (a * a - b * b) / (b * b)
  A <- 1 + uSq / 16384 * (4096 + uSq * (-768 + uSq * (320 - 175 * uSq)))
  B <- uSq/1024 * (256 + uSq * (-128 + uSq * (74 - 47 * uSq)))
  deltaSigma <- B * sinSigma *
                 (cos2SigmaM +
                  B / 4 * (cosSigma * (-1 + 2 * cos2SigmaM^2) -
                           B / 6 * cos2SigmaM *
                           (-3 + 4 * sinSigma^2) *
                           (-3 + 4 * cos2SigmaM^2)
                          )
                 )
  s <- b * A * (sigma - deltaSigma)
  return(s) # Distance in km
}

# Create additional function that incorporates the greatCircleDistance function.
protectedGreatCircleDistance <- function(lat1, lon1, lat2, lon2) {
        if(any(is.na(c(lat1, lon1, lat2, lon2)))) { 
            rc <- NA
        } else {
            rc <- greatCircleDistance(lat1, lon1, lat2, lon2)
        }
        return(rc)
    }
```

Once we have that function, we can start to prepare the data we need to determine transect spacing. For boatable sites, we start with the `nrsaChanrip` dataset and extract the LATDD_TOP and LONDD_TOP parameters, which are the coordinates at each transect. The coordinates were measured at the bank and mid-stream, and the mid-stream ones are preferred, but bank coordinates are used if those are not available for a given transect for some reason.

We need to then create a column for each transect that will eventually be filled with the distance from that transect to the next for the actual transect spacing. The `greatCircleDistance` will calculate the distance between transects based on latitude/longitude, but in case there are errors in coordinates, there is a second set of calculations to fill in the median where the distance appears to be an error.

```{r nrsaGeneral-boatable-spacing}
bLatLon <- nrsaChanrip |>
  filter(PARAMETER %in% c('LATDD_TOP', 'LONDD_TOP')) |>
  mutate(BANK = ifelse(BANK=='BANK', 'BANK', 'MID'),
         PARAMETER = ifelse(PARAMETER=='LATDD_TOP', 'LAT', 'LON'),
         BANKPARAM = paste(BANK, PARAMETER, sep='_')) |>
  pivot_wider(id_cols = c('SITE', 'TRANSECT'), names_from=BANKPARAM, values_from='VALUE') |>
  mutate(LAT = ifelse(is.na(MID_LAT)|is.na(MID_LON), as.numeric(BANK_LAT), as.numeric(MID_LAT)),
         LON = ifelse(is.na(MID_LON)|is.na(MID_LON), as.numeric(BANK_LON), as.numeric(MID_LON)))|>
  select(-BANK_LAT, -BANK_LON, -MID_LAT, -MID_LON) |>
  pivot_wider(id_cols = 'SITE', names_from='TRANSECT', values_from=c('LAT', 'LON')) |>
  within(A<-B<-C<-D<-E<-F<-G<-H<-I<-J<-as.numeric(NA)) |> # Creates empty columns with each transect name, in order
  ddply(c('SITE'), summarise, 
         A = protectedGreatCircleDistance(LAT_A, LON_A, LAT_B, LON_B),
         B = protectedGreatCircleDistance(LAT_B, LON_B, LAT_C, LON_C),
         C = protectedGreatCircleDistance(LAT_C, LON_C, LAT_D, LON_D),
         D = protectedGreatCircleDistance(LAT_D, LON_D, LAT_E, LON_E),
         E = protectedGreatCircleDistance(LAT_E, LON_E, LAT_F, LON_F),
         F = protectedGreatCircleDistance(LAT_F, LON_F, LAT_G, LON_G),
         G = protectedGreatCircleDistance(LAT_G, LON_G, LAT_H, LON_H),
         H = protectedGreatCircleDistance(LAT_H, LON_H, LAT_I, LON_I),
         I = protectedGreatCircleDistance(LAT_I, LON_I, LAT_J, LON_J),
         J = protectedGreatCircleDistance(LAT_J, LON_J, LAT_K, LON_K))

bTransectSpacing <- bLatLon |>
  pivot_longer(cols = A:J, names_to = 'TRANSECT', values_to='DISTANCE') |>
  ddply('SITE', mutate, medianDistance = median(DISTANCE, na.rm=T)) |>
  mutate(correctedDistance = ifelse(abs((DISTANCE-medianDistance)/medianDistance) >0.5, 
                                    medianDistance, DISTANCE)) |>
  dplyr::rename(c(VALUE = correctedDistance)) |>
  select(SITE, TRANSECT, VALUE)


```

#### Wadeable transect spacing

We now have transect spacing for boatable sites but need it for wadeable sites, too. Instead of using coordinates, we can just use thalweg data (`nrsaThalweg`). Since all wadeable sites have INCREMENT filled in our dataset, we can use that for transect spacing and do not need to worry about filling in any alternatives. If any wadeable sites did not have the INCREMENT parameter, we would need to find an alternative value, which could be based on the reach length recorded on the thalweg form or the verification form and divided by the number of subreaches sampled. The increment indicates the intended spacing of stations within a subreach between transects.

We do have to figure out how many stations were sampled, specifically at the last transect, so that we can adjust the total number of stations. There are a few steps for doing this provided below. Essentially, we count the number of stations sampled in a subreach (between transects). We need to use the internal `aquamet` function `nWadeableStationsPerTransect()`, which essentially determines the number of stations within each subreach using counts of stations with measurements.

```{r nrsaGeneral-wadeable-spacing}
# Pull increment, which is only measured at wadeable sites
wThalwegIncrements <- nrsaThalweg |>
  filter(PARAMETER=='INCREMENT') |>
  mutate(VALUE = as.numeric(VALUE)) |>
  select(SITE, VALUE)

# We still need to create a data frame of the stations sampled and transect distance
# We can get the transects sampled using the thalweg data
wNumStations <- nrsaThalweg |>
  filter(STATION!='ALL') |>
  mutate(STATION = as.integer(STATION)) |>
  aquamet::nWadeableStationsPerTransect()
  
wNumStationsPerTransect <- filter(nrsaThalweg, VALUE!='' & !is.na(VALUE) & TRANSECT %in% LETTERS[1:11]) |>
  ddply('SITE', summarise, lastTransect = max(TRANSECT)) |>
  merge(wNumStations, by = 'SITE', all.y=T) |>
  mutate(nSta = ifelse(TRANSECT==lastTransect, nSta-1, nSta)) |>
  select(-lastTransect)

wTransectSpacing <- merge(wThalwegIncrements, wNumStationsPerTransect, by = 'SITE') |>
  mutate(VALUE = VALUE * nSta) |>
  select(SITE, TRANSECT, VALUE) |>
  arrange(SITE, TRANSECT)

```

#### Calculating general metrics

We are finally ready to use the `nrsaGeneral()` function to calculate metrics. There are a couple of additional inputs which are subsets of the `nrsaThalweg` data. Specifically, we need the set of sampled transects and we need any data from side channels.

```{r nrsaGeneral-metrics}
sampTrans <- nrsaThalweg |>
  select(SITE, TRANSECT) |>
  unique()

sideChan <- nrsaThalweg |>
  filter(PARAMETER %in% c('SIDCHN', 'OFF_CHAN')) |>
  select(SITE, VALUE)
  
transSpacing <- bind_rows(bTransectSpacing, wTransectSpacing)

generalMets <- nrsaGeneral(sampledTransects = sampTrans, 
                           sideChannels = sideChan,
                           transectSpacing = transSpacing)

print("The general site metrics from this function:")
unique(generalMets$METRIC)
```

### NRSA human influence metrics

There are a number of metrics based on human influence data collected on the **PHAB CHANNEL/RIPARIAN TRANSECT** form for boatable sites and the **PHAB CHANNEL/RIPARIAN CROSS-SECTION** form for wadeable sites. These data are included in the `nrsaChanrip` dataset. The parameters have the SAMPLE_TYPE value of HUMINFLUW or HUMINFLUB, for wadeable and boatable sites, respectively.

For this function, we simply provide inputs for each type of human influence as a data frame with SITE and VALUE. We can wrap all of this subsetting into the function call, so the inputs are not created a separate datasets. We also need to be sure we only have the following variables in the data frame: SITE, TRANSECT, VALUE

```{r nrsaHumanInfluence}
nrsaHumInflMets <- nrsaHumanInfluence(
  buildings = filter(nrsaChanrip, PARAMETER == 'BUILD') |> 
    select(SITE, TRANSECT, VALUE),
  landfillTrash = filter(nrsaChanrip, PARAMETER == 'TRASH') |> 
    select(SITE, TRANSECT, VALUE),
  logging = filter(nrsaChanrip, PARAMETER == 'LOG') |> 
    select(SITE, TRANSECT, VALUE),
  mining = filter(nrsaChanrip, PARAMETER == 'MINE') |> 
    select(SITE, TRANSECT, VALUE),
  parkLawn = filter(nrsaChanrip, PARAMETER == 'PARK') |> 
    select(SITE, TRANSECT, VALUE),
  pastureRangeHay = filter(nrsaChanrip, PARAMETER == 'PAST') |> 
    select(SITE, TRANSECT, VALUE),
  pavementClearedlot = filter(nrsaChanrip, PARAMETER == 'PAVE') |> 
    select(SITE, TRANSECT, VALUE),
  pipesInOut = filter(nrsaChanrip, PARAMETER == 'PIPES') |> 
    select(SITE, TRANSECT, VALUE),
  roadsRailroads = filter(nrsaChanrip, PARAMETER == 'ROAD') |> 
    select(SITE, TRANSECT, VALUE),
  rowcrops = filter(nrsaChanrip, PARAMETER == 'ROW') |> 
    select(SITE, TRANSECT, VALUE),
  wallRevetment = filter(nrsaChanrip, PARAMETER == 'WALL') |> 
    select(SITE, TRANSECT, VALUE))

print("The human influence metrics from this function:")
unique(nrsaHumInflMets$METRIC)
```

### NRSA large woody debris metrics

The large woody debris metrics are calculated using the function `nrsaLargeWoody()` and based on data collected at each transect both in and out of the water. The data for this example are in the `nrsaChannel` dataset with the SAMPLE_TYPE values of LWDW and LWDB for wadeable and boatable sites, respectively. The inputs are separated into boatable and wadeable for the function. In addition, certain metrics already calculated are used as inputs. The `reachlen` metric from the `nrsaGeneral()` function (`generalMets` dataset) and the `xbkf_w` metric from the `nrsaChannelMorphology()` function (`chanMorphMets` dataset) are both used for the large woody debris metric calculations. For the purposes of this function, the `PARAMETER` should be renamed to `CLASS` for the `bCounts` and `wCounts` inputs.

```{r nrsaLargeWoody}
lwd <- filter(nrsaChannel, SAMPLE_TYPE %in% c('LWDB', 'LWDW')) |>
  mutate(VALUE = as.integer(VALUE))

largeWoodyDebrisMets <- nrsaLargeWoody(
  bCounts = filter(lwd, SAMPLE_TYPE=='LWDB') |>
    select(SITE, TRANSECT, CLASS = PARAMETER, VALUE),
  wCounts = filter(lwd, SAMPLE_TYPE=='LWDW') |>
    select(SITE, TRANSECT, CLASS = PARAMETER, VALUE),
  reachlength = filter(generalMets, METRIC=='reachlen') |>
    select(SITE, VALUE),
  meanBankfullWidth = filter(chanMorphMets, METRIC=='xbkf_w') |>
    select(SITE, VALUE)
                    )

print("The large woody debris metrics from this function:")
unique(largeWoodyDebrisMets$METRIC)
```

### NRSA littoral depth metrics

Littoral depth metrics only apply to boatable sites because those are the only sites where depth sampling along the shoreline is measured. The measurements are recorded on the **PHAB CHANNEL/RIPARIAN TRANSECT** form for boatable sites and the data for this example is in the dataset `nrsaLittoral`. All littoral measurements should be converted to meters before using in the function.

```{r nrsaLittoralDepth}
units <- nrsaLittoral |>
  filter(PARAMETER=='DEPTH_UNITS') |>
  select(SITE, TRANSECT, UNITS=VALUE)

depths <- nrsaLittoral |>
  filter(PARAMETER %in% c('SONAR', 'POLE')) |>
  mutate(VALUE = as.numeric(VALUE)) |>
  merge(units, by = c('SITE', 'TRANSECT')) |>
  select(SITE, VALUE)

littoralDepthMets <- nrsaLittoralDepth(bLittoralDepth = depths)

print("The littoral depth metrics from this function:")
unique(littoralDepthMets$METRIC)
```

### NRSA riparian vegetation metrics

Riparian vegetation data are collected at each transect, and the data for this example are contained in the `nrsaChanrip` dataset. The function `nrsaRiparianVegetation()` is used to calculate these metrics, and the inputs are fairly straightforward, just the data for individual parameters.  

```{r nrsaRiparianVegetation}
nrsaRiparianVegMetrics <- nrsaRiparianVegetation(
  canopyCoverLargeDiameter = filter(nrsaChanrip, PARAMETER=='CANBTRE') |>
    select(SITE, TRANSECT, BANK, VALUE),
  canopyCoverSmallDiameter = filter(nrsaChanrip, PARAMETER=='CANSTRE') |>
    select(SITE, TRANSECT, BANK, VALUE),
  canopyVegetationType = filter(nrsaChanrip, PARAMETER=='CANVEG') |>
    select(SITE, TRANSECT, BANK, VALUE),
  groundCoverBare = filter(nrsaChanrip, PARAMETER=='BARE') |>
    select(SITE, TRANSECT, BANK, VALUE),
  groundCoverNonwoody = filter(nrsaChanrip, PARAMETER=='GCNWDY') |>
    select(SITE, TRANSECT, BANK, VALUE),
  groundCoverWoody = filter(nrsaChanrip, PARAMETER=='GCWDY') |>
    select(SITE, TRANSECT, BANK, VALUE),
  understoryCoverNonwoody = filter(nrsaChanrip, PARAMETER=='UNDNWDY') |>
    select(SITE, TRANSECT, BANK, VALUE),
  understoryCoverWoody = filter(nrsaChanrip, PARAMETER=='UNDWDY') |>
    select(SITE, TRANSECT, BANK, VALUE),
  understoryVegetationType = filter(nrsaChanrip, PARAMETER=='UNDERVEG') |>
    select(SITE, TRANSECT, BANK, VALUE)
)

print("The riparian vegetation metrics from this function:")
unique(nrsaRiparianVegMetrics$METRIC)
```

### NRSA slope and bearing metrics

Slope and bearing measurements are recorded in the field only for wadeable sites, and slopes for boatable sites are usually generated using GIS. In this example dataset from data collected in the 2008-9 NRSA survey, we have slope and bearing measurements for all sites. The inputs for the function `nrsaSlopeBearing()` are separated into boatable and wadeable, and if sinuosity and slope were determined through GIS for any sites, those are added through a separate set of arguments. 

For demonstration purposes, we will go through an example for boatable sites using field-collected data, even though we no longer collect data that way. We will create the inputs for boatable sites separately because those data are more complex than for wadeable sites. 

```{r nrsaSlopeBearing}
boatBearing <- filter(nrsaSlope, SAMPLE_TYPE=='BANKB')

slopeBearingMets <- nrsaSlopeBearing(
  bBearing = filter(nrsaSlope, PARAMETER==')
)
```

